---
title: H5N1 outbreak scenarios
---

This report requires two R packages: {epichains} and {ggplot2}, the former to simulate the branching process model and the latter to visualise the results.

```{r echo=FALSE}
# install from outbreak_dist branch of {epichains}
# remotes::install_github("epiverse-trace/epichains@outbreak_dist")
library(epichains)
library(ggplot2)
```

The `simulate_scenarios()` function from {epichains} calculates the proportion of outbreak statistic (size and/or length) across scenarios of reproduction number ($R0$) values, overdispersion, $k$, and offspring distributions.

Here, we use a simple branching process to explore scenarios of reproduction number ($R0$) values, overdispersion, $k$, and offspring distributions and their associated chain size and length. The branching process is seeded with an initial number of cases and an offspring distribution, and optionally, a generation time distribution.

We define the chain size as the total number of cases produced by a chain before it goes extinct. The chain length, on the other hand, is the total number of generations reached by a chain before it goes extinct.

We use a reproduction number ($R$) between 0.1 and 0.9 at 0.1 intervals, and for the Negative Binomial offspring distribution, we vary the dispersion parameter ($k$) between 0.5 and 2.5 at 0.5 intervals. Smaller values of $k$ signify more overdispersion in the individual-level transmissibility, in other words transmission is more heterogeneous and superspreading events are more likely. As $k$ increases individual-level transmission is more homogeneous and when $k \rightarrow \infty$ the Negative Binomial equals the Poisson offspring distribution. 

```{r}
df <- simulate_scenarios(
  offspring_dist = c("rpois", "rnbinom"), 
  statistic = c("size", "length"), 
  R_seq = seq(0, 0.9, 0.1), 
  k_seq = seq(0.5, 2.5, 0.5), 
  breaks = c(0, 2, 5, 10, 20, 50, 100, Inf)
)
```

The function produces a `data.frame` that contains the outbreak size or length in intervals (`$interval`), the proportion of outbreaks that fall into that interval (`$Freq`), the reproduction number (`$R`), the dispersion parameter (`$k`, for Poisson distributions $k$ is set to `NA`), the offspring distribution (`$offspring_dist`), and the outbreak statistic (`$statistic`, either size or length). 

Here are the first 6 rows.

```{r}
head(df)
```

Below, we show the results for the Poisson distribution for varying plausible $R0$ values. The figure shows the proportion of outbreaks that achieved different outbreak size ranges.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
df |> 
  subset(offspring_dist == "rpois") |>
  ggplot2::ggplot() +
  ggplot2::geom_col(
    mapping = ggplot2::aes(x = as.factor(R), y = Freq, fill = interval)
  ) +
  ggplot2::scale_x_discrete(name = "Reproduction number (R)") +
  ggplot2::scale_y_continuous(name = "Proportion of outbreaks") +
  ggplot2::scale_fill_brewer(name = "Outbreak size", palette = "Spectral") +
  ggplot2::facet_wrap(
    facets = c("offspring_dist", "statistic"), 
    labeller = ggplot2::label_both
  ) +
  ggplot2::labs(
    title = "Outbreak size and length for Poisson Branching Process"
  ) +
  ggplot2::theme_bw() + 
  ggplot2::theme(strip.background = ggplot2::element_blank())
```

Next are the results of the negative distribution offspring for varying plausible $R0$ values but a fixed $k$ of $0.5$. The figure shows the proportion of outbreaks that achieved different outbreak size ranges.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
df |> 
  subset(offspring_dist == "rnbinom" & k == 0.5) |>
  ggplot2::ggplot() +
  ggplot2::geom_col(
    mapping = ggplot2::aes(x = as.factor(R), y = Freq, fill = interval)
  ) +
  ggplot2::scale_x_discrete(name = "Reproduction number (R)") +
  ggplot2::scale_y_continuous(name = "Proportion of outbreaks") +
  ggplot2::scale_fill_brewer(name = "Outbreak size", palette = "Spectral") +
  ggplot2::facet_wrap(
    facets = c("offspring_dist", "statistic"), 
    labeller = ggplot2::label_both
  ) +
  ggplot2::labs(
    title = "Outbreak size and length for Negative Binomial Branching Process"
  ) +
  ggplot2::theme_bw() + 
  ggplot2::theme(strip.background = ggplot2::element_blank())
```

We now show the results of the outbreak using a negative distribution offspring for varying plausible values of $R0$ and $k$. The figure shows the proportion of outbreaks that achieved different outbreak size ranges.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
df |> 
  subset(offspring_dist == "rnbinom" & statistic == "size") |>
  ggplot2::ggplot() +
  ggplot2::geom_col(
    mapping = ggplot2::aes(x = as.factor(R), y = Freq, fill = interval)
  ) +
  ggplot2::scale_x_discrete(name = "Reproduction number (R)") +
  ggplot2::scale_y_continuous(name = "Proportion of outbreaks") +
  ggplot2::scale_fill_brewer(name = "Outbreak size", palette = "Spectral") +
  ggplot2::facet_wrap(
    facets = c("k"), 
    labeller = ggplot2::label_both
  ) +
  ggplot2::labs(
    title = "Outbreak size Negative Binomial Branching Process"
  ) +
  ggplot2::theme_bw() + 
  ggplot2::theme(strip.background = ggplot2::element_blank())
```

In the next figure, we show the results of outbreak lengths for a negative binomial offspring distribution with varying $R0$ and $k$ values.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
df |> 
  subset(offspring_dist == "rnbinom" & statistic == "length") |>
  ggplot2::ggplot() +
  ggplot2::geom_col(
    mapping = ggplot2::aes(x = as.factor(R), y = Freq, fill = interval)
  ) +
  ggplot2::scale_x_discrete(name = "Reproduction number (R)") +
  ggplot2::scale_y_continuous(name = "Proportion of outbreaks") +
  ggplot2::scale_fill_brewer(name = "Outbreak length", palette = "Spectral") +
  ggplot2::facet_wrap(
    facets = c("k"), 
    labeller = ggplot2::label_both
  ) +
  ggplot2::labs(
    title = "Outbreak length for Negative Binomial Branching Process"
  ) +
  ggplot2::theme_bw() + 
  ggplot2::theme(strip.background = ggplot2::element_blank())
```
