---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Modelling scenarios of importation of H5N1 in the UK and the impact of interventions

<!-- badges: start -->
<!-- badges: end -->

This repository contains the code to reproduce the analyses of: [Ward et al. (2024) Estimates of epidemiological parameters for H5N1 influenza in humans: a rapid review. https://doi.org/10.1101/2024.12.11.24318702.](https://doi.org/10.1101/2024.12.11.24318702).

## Package structure

The package is organised into analysis scripts in the `scripts/` folder, the plots included in the paper are in the `plots/` folder. 

```{r}
library(fs)
fs::dir_tree()
```

## Analyses

### Outbreak size distribution

The script to run the outbreak size and length distribution simulation and generate the plots is in `scripts/outbreak_distribution.R`. The plots are included in the `plot` folder, including the The [`outbreak_size` plot](plots/outbreak_size.png) and [`outbreak_length` plot](plots/outbreak_length.png).
